---
title: "Incarceration Rates by Race"
author: "Sarah Schafer"
date: "11/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(mapproj)
source("analysis.R")
```
## Introduction
Note: While pie charts can be difficult to interpret, in this situation, where I wanted to display how the racial demographics of jailed populations differs from the racial demographics of the general population.  
``` {r echo = FALSE}
ggarrange(pie_jail, pie_total, ncol = 2, nrow = 2)

```

## Comparison over time charts
``` {r echo = FALSE}
ggarrange(jail_percent_time_plot, total_percent_time_plot, total_jail_time_plot, total_time_plot, ncol = 2, nrow = 2)
```

## Variable Comparison Charts
``` {r echo = FALSE}
ggarrange(total_jail_to_black_pop, total_jail_to_white_pop, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
```


## Maps
``` {r echo = FALSE}
black_pop_map
jail_pop_map
```

## Sources
The data I used was from the Vera Institute of Justice, which has a [data set](https://github.com/vera-institute/incarceration-trends/blob/master/incarceration_trends.csv) containing a large amount of data on incarceration demographics, disaggregated by race, as well as gender.  

For this project, I researched to find additional packages and ways to create visualizations to accurately display the data beyond the methods we discussed in class.  Here are some of the references I used in order to create this project:
