---
title: "Incarceration Rates by Race"
author: "Sarah Schafer"
date: "11/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(mapproj)
source("analysis.R")
```
## Introduction
Note: While pie charts can be difficult to interpret, in this situation, where I wanted to display how the composition of the jailed population  differs from the composition of the general population, it was the best visualization I could create.  The general population only shows ages 15-64, as this was the data included in the data set.  
``` {r pie_charts, echo = FALSE, message = FALSE}
ggarrange(pie_jail, pie_total, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
```

## Comparison over time charts
``` {r time_charts, echo = FALSE, message = FALSE}
ggarrange(jail_percent_time_plot, total_percent_time_plot, total_jail_time_plot, total_time_plot, ncol = 2, nrow = 2, common.legend = TRUE, legend = "right")
```

## Variable Comparison Charts
``` {r var_comp_charts, echo = FALSE, message = FALSE}
ggarrange(total_jail_to_black_pop, total_jail_to_white_pop, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
```


## Maps
``` {r map_charts, echo = FALSE, message = FALSE, warning = FALSE}
black_pop_map
jail_pop_map
```

## Sources
The data I used was from the Vera Institute of Justice, which has a [data set](https://github.com/vera-institute/incarceration-trends/blob/master/incarceration_trends.csv) containing a large amount of data on incarceration demographics, disaggregated by race, as well as gender.  

For this project, I researched to find additional packages and ways to create visualizations to accurately display the data beyond the methods we discussed in class.  Here are some of the references I used in order to create this project:

- To make the pie charts, I referenced [the R Graph Gallery](https://www.r-graph-gallery.com/piechart-ggplot2.html).

- To get rid of the warning messages that appeared in my Rmd file, I used this solution from [Stack Overflow](https://stackoverflow.com/questions/45399587/how-to-remove-warning-messages-in-r-markdown-document).
